---
title: "DC Youth Stats"
output:
  html_document:
    code_folding: hide
date: "2026-08-12"
---

<style>
  body { font-size: 18px; }
  p { font-size: 18px; }
</style>



```{r setup, echo = T, warning=F, message=F, }
knitr::opts_chunk$set(echo = T, warning=F, message=F, out.width = "100%")

library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(stringr)
library(RColorBrewer)
library(sf)
library(tidycensus) 
library(purrr) 
library(patchwork)
library(plotly)
library(readxl)
library(srvyr)


source("api_keys.R")

setwd("~/GitHub/youth-stats")
ms=1.1
total_col="steelblue"; black_col="#8d5524"; white_col="#ffdbac"
hispa_col="#e0ac69"; other_col="#702963"

fmt <- function(num) {
  format(round(num), big.mark=",")
}
```


<br>

Many U.S. cities have seen their youth-age populations decline. See [this WSJ piece](https://www.wsj.com/us-news/cities-losing-families-children-a860a9b1?st=w78QhH&reflink=desktopwebshare_permalink), for example. 

DC is an exception. 

This page looks at why that might be.


# How many kids live in DC?

The chart below shows DC's youth population between 1990 and 2025. After falling between 1995 and 2010, the number has ticked up again, with a COVID-induced dip recently.

<br>

```{r}
cpop <- read_excel("data/Child population by gender.xlsx") %>%
  filter(Gender == "Total less than 18", DataFormat=="Number") %>%
  mutate(count = as.numeric(Data), year=as.numeric(TimeFrame))

ggplot(cpop, aes(x=year, y=as.numeric(count))) +
  geom_line(linewidth = ms) +
   scale_y_continuous(
      labels = scales::label_comma()
    , limits = c(0, 140e3)
    ) +
  labs(x="", y="Number of Children", 
       caption="Chart shows population under 18. Source: KidsCount / The Annie E. Casey Foundation, U.S. Census Bureau",
       title = "After falling from '95-'10, DC's youth population rose"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    # , panel.grid = element_blank()
    )

```


<br>

If we look farther back in time, we can see that DC's youth population used to be much higher, both in absolute and percentage terms. The percentage of children has continued to fall, even as the number has risen. (Note that this data uses "under 15" as opposed to "under 18.")

<br>

```{r}
df <- read_excel("data/OP_pop_data.xlsx")

a <- 
  ggplot(df, aes(x=year, y=u15_pop)) +
  geom_line(linewidth = ms) +
  scale_y_continuous(labels = scales::label_comma(), limits = c(0, 200e3)) +
  scale_x_continuous(n.breaks = 10) +
  theme_minimal() +
    theme(
     plot.caption = element_text(hjust = 0)
    # , panel.grid = element_blank()
    ) +
  labs(x="", y="Under 15 population")

b <- 
  ggplot(df, aes(x=year, y=u15_pct)) +
  geom_line(linewidth = ms) +
  scale_y_continuous(labels = scales::label_percent(), limits=c(0, .3)) +
  scale_x_continuous(n.breaks = 10) +
  theme_minimal() +
    theme(
     plot.caption = element_text(hjust = 0)
    # , panel.grid = element_blank()
    ) +
  labs(x="", y="Under 15 population, as % of total pop.")

a + b + plot_annotation(
  title = "DC's youth population used to be much higher",
  caption = 'Source: DC Office of Planning, U.S. Census Bureau',
  theme = theme(
      plot.title = element_text(size = 16, face = 'bold')
    , plot.caption = element_text(hjust = 0)
    )
  )
```



```{r}

sfx <- c("", "A","B","C","D","E","F","G","H","I")
lbl <- c("Total, all races","White alone","Black alone","AIAN alone","Asian alone",
         "NHPI alone","Some other race alone","Two or more races",
         "Hispanic or Latino","White alone, not Hispanic")

kid_cells <- c(3:6, 27:30)   # M/F: under 5, 5-9, 10-14, 15-17

vfmt <- function(year, s, i)
if (year == 2020) sprintf("P12%s_%03dN", s, i) else sprintf("P012%s%03d", s, i)

pull_kids <- function(year, sumfile, geography, ...) {
  map2_dfr(sfx, lbl, function(s, l) {
    vars <- setNames(vfmt(year, s, c(1, kid_cells)),
                     c("pop_total", paste0("c", kid_cells)))
    get_decennial(geography = geography, variables = vars, year = year, key = CENSUS_KEY,
                  sumfile = sumfile, output = "wide", ...) |>
      mutate(race = l, year = year)
  }) |>
    mutate(children = rowSums(across(num_range("c", kid_cells)))) |>
    transmute(GEOID, NAME, year, race,
              population = pop_total,
              children,
              share_children = children / pop_total)
}


vint <- tibble::tibble(year = c(2000, 2010, 2020), sumfile = c("sf1","sf1","dhc"))
all_counties <- map2_dfr(vint$year, vint$sumfile, ~pull_kids(.x, .y, "county"))

# chk <- get_decennial("county", c(p1 = "P1_001N"), year = 2020, sumfile = "dhc", key = CENSUS_KEY)
# all_counties |>
#   filter(year == 2020, race=="Total, all races") |>
#   inner_join(chk, by="GEOID") |>
#   summarise(max_diff = max(abs(population - value)))

xwalk <- tibble::tribble(
  ~city,            ~state, ~county,
  "New York",       "NY",  "Bronx County",
  "New York",       "NY",  "Kings County",
  "New York",       "NY",  "New York County",
  "New York",       "NY",  "Queens County",
  "New York",       "NY",  "Richmond County",
  "Los Angeles",    "CA",  "Los Angeles County",
  "Chicago",        "IL",  "Cook County",
  "Houston",        "TX",  "Harris County",
  "Phoenix",        "AZ",  "Maricopa County",
  "Philadelphia",   "PA",  "Philadelphia County",
  "San Antonio",    "TX",  "Bexar County",
  "San Diego",      "CA",  "San Diego County",
  "Dallas",         "TX",  "Dallas County",
  "Jacksonville",   "FL",  "Duval County",
  "Austin",         "TX",  "Travis County",
  "Fort Worth",     "TX",  "Tarrant County",
  "San Jose",       "CA",  "Santa Clara County",
  "Columbus",       "OH",  "Franklin County",
  "Charlotte",      "NC",  "Mecklenburg County",
  "Indianapolis",   "IN",  "Marion County",
  "San Francisco",  "CA",  "San Francisco County",
  "Seattle",        "WA",  "King County",
  "Denver",         "CO",  "Denver County",
  "Oklahoma City",  "OK",  "Oklahoma County",
  "Nashville",      "TN",  "Davidson County",
  "Washington",     "DC",  "District of Columbia",
  "El Paso",        "TX",  "El Paso County",
  "Las Vegas",      "NV",  "Clark County",
  "Boston",         "MA",  "Suffolk County",
  "Detroit",        "MI",  "Wayne County",
  "Portland",       "OR",  "Multnomah County",
  "Louisville",     "KY",  "Jefferson County",
  "Memphis",        "TN",  "Shelby County",
  "Baltimore",      "MD",  "Baltimore city",
  "Milwaukee",      "WI",  "Milwaukee County",
  "Albuquerque",    "NM",  "Bernalillo County",
  "Tucson",         "AZ",  "Pima County",
  "Fresno",         "CA",  "Fresno County",
  "Sacramento",     "CA",  "Sacramento County") |>
  left_join(distinct(transmute(fips_codes, state, county,
                               GEOID = paste0(state_code, county_code))),
            by = c("state", "county"))

stopifnot(!any(is.na(xwalk$GEOID)))   # catches any name I got wrong

out_c <- all_counties |>
  inner_join(xwalk, by = "GEOID") |>
  arrange(city, year, race)
```

<br>

Relative to other U.S. urban counties, DC has relatively few children, both in absolute terms, and as a share of its population. (Keep in mind that the data below pulls information at the county-level, so DC suburbs are excluded, which may make the counts look somewhat lower than in, say, Houston, where Harris County encompasses some of the city's suburbs).

<br>

```{r}
a <- out_c %>% filter(race == "Total, all races", year==2020) %>%
  group_by(city) %>%
  summarise(children=sum(children, na.rm=T), .groups='drop') %>%
  mutate(DC = if_else(city == "Washington", 1, 0)) %>%
  ggplot(aes(y=forcats::fct_reorder(city, children), x=children/1000, fill=factor(DC))) +
  geom_bar(stat="identity") +
  labs(x="Number of children (thousands)", y="") + 
  guides(fill="none") +
  scale_x_continuous(labels = scales::label_comma())

b <- out_c %>% filter(race == "Total, all races", year==2020) %>%
  group_by(city) %>%
  summarise(
      children=sum(children, na.rm=T)
    , population=sum(population, na.rm=T)
    , share_children = children / population
    , .groups='drop') %>%
  mutate(DC = if_else(city == "Washington", 1, 0)) %>%
  ggplot(aes(y=forcats::fct_reorder(city, share_children), x=share_children, fill=factor(DC))) +
  geom_bar(stat="identity") +
  labs(x="Children as share of population", y="") + 
  guides(fill="none") +
  scale_x_continuous(labels = scales::label_percent())

a + b + plot_annotation(
  title = 'DC has relatively few children',
  caption = 'Charts show county-level data for large U.S. cities in 2020. Source: U.S. Census Bureau, 2020 Decennial Census.',
  theme = theme(
      plot.title = element_text(size = 16, face = 'bold')
    , plot.caption = element_text(hjust = 0)
    )
  )
```

<br>

While some other cities have had longer-term increases in the number of children, DC's increase has been more of a return to a level previously seen in the 1990s. The line charts below show number of children in 2000, 2010, and 2020.

<br>

```{r}
  out_c %>% filter(race == "Total, all races") %>%
  group_by(year, city) %>% summarise(children = sum(children, na.rm=T), .groups='drop') %>%
  group_by(city) %>%
  mutate(rise = if_else(children[year==2010] < children[year==2020], 1, 0)) %>% ungroup() %>%
  # mutate(rise = if_else(city == "Washington", 2, rise)) %>%
  ggplot(aes(x=year, y=children, color=factor(rise))) +
  geom_line() +
  labs(x="", y="Number of children") +
  guides(color="none") +
  scale_x_continuous(n.breaks = 3) +
  scale_color_manual(values = c("#FF3F00", "#70B77E")) +
  theme(
    panel.grid = element_blank(),
    axis.title.y = element_blank(), # Removes the main Y-axis title ("mpg")
    axis.text.y  = element_blank(), # Removes the tick labels (numbers)
    axis.ticks.y = element_blank(), # Removes the small tick marks
    
    axis.ticks.x = element_blank(),  # Removes the small tick marks
    axis.text.x  = element_blank()  # Removes the tick labels (numbers)
  ) +
  facet_wrap(~forcats::fct_reorder(city, rise), scales="free_y") +
  ggtitle("DC child population changes compared to other places")
```



```{r}
plt <-
  ggplot(out_c %>% filter(city=="Washington", race != "White alone, not Hispanic") %>%
           group_by(city, year) %>%
           mutate(
             race = case_when(
               race %in% c(
               "AIAN alone", "Asian alone", "NHPI alone", 
               "Some other race alone", "Two or more races"
             ) ~ "Other"
             , race == "Black alone" ~ "Black"
             , race == "Hispanic or Latino" ~ "Hispanic"
             , race == "White alone" ~ "White"
             , race == "Total, all races" ~ "All"
             , .default = race
             )
           ) %>% ungroup() %>%
           group_by(city, year, race) %>%
           summarise(children = sum(children), .groups  = 'drop')
         , aes(x=year, y=children, color=factor(race))
         ) +
  geom_line(linewidth=ms) +
  scale_y_continuous(
      labels = scales::label_comma()
    # , limits = c(0, 115e3)
    ) +
  scale_color_manual(values = c(
              "All"=total_col
            , "Black"=black_col
            , "Hispanic"=hispa_col
            , "Other"=other_col
            , "White"=white_col
            )) +
  labs(x="", y="Number of children", color="",
       caption="Hispanic category not mutually exclusive with other categories.\nSource: U.S. Census Bureau, Decennial Census.",
       title = "Children in DC by race: fewer Black children, other groups saw increase"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    # , panel.grid = element_blank()
    )
```

<br>

While the numbers of Hispanic and White children have risen in DC since 2010, the number of Black children has continued to fall. 

<br>


```{r}
plt
```






```{r}
clean_data <- function(df) {
  df %>%
  pivot_longer(
    cols = -c(  `School Name` 
              , `State Name [Public School] Latest available year`
              , `School ID (12-digit) - NCES Assigned [Public School] Latest available year`
              ), 
    names_to = c(".value", "school_year"),
    names_pattern = "(.*)(.{7})$"
  ) %>%
    rename(
        "id" = "School ID (12-digit) - NCES Assigned [Public School] Latest available year"
      , "name" = "School Name"
      , "state" = "State Name [Public School] Latest available year"   
      ) %>%
    rename_with(~ str_remove(., fixed("[Public School]")), everything()) %>%
    rename_with(str_trim) %>%
    filter(!is.na(state)) %>%
    select(-state)
}






# Data from here:
# https://nces.ed.gov/ccd/elsi/tableGenerator.aspx


df1 <- read_csv("~/GitHub/youth-stats/data/ELSI_csv_export_6392212884336770323709.csv", skip = 5) %>% 
  clean_data(.) %>%
  rename(lat=Latitude, lon=Longitude)

df2 <- read_csv("~/GitHub/youth-stats/data/ELSI_csv_export_6392212902055494018674.csv", skip = 5) %>% 
  clean_data(.)

df3 <- read_csv("data/ELSI_csv_export_6392212906851656167285.csv", skip = 5) %>% 
  clean_data(.) %>%
  select(-`Grades 1-8 Students`, -`Grades 9-12 Students`)

df4 <- read_csv("data/ELSI_csv_export_639221340086678151282.csv", skip=5) %>%
  clean_data(.)

df5 <- read_csv("data/ELSI_csv_export_6392213789371741314818.csv", skip=5) %>%
  clean_data(.)

df <-
  dplyr::full_join(df1, df2, by = c("id", "name", "school_year")) %>%
  dplyr::full_join(df3, by = c("id", "name", "school_year")) %>%
  dplyr::full_join(df4, by = c("id", "name", "school_year")) %>%
  dplyr::full_join(df5, by = c("id", "name", "school_year")) %>%
  mutate(id = as.character(id)) %>%
  mutate(across(everything(), ~ na_if(., "†"))) %>%
  mutate(across(everything(), ~ na_if(., "–"))) %>%
  mutate(across(everything(), ~ na_if(., "‡"))) %>%
  rename(addr = `Location Address 1`) %>%
  group_by(id, name) %>%
  mutate(
      lat = first(lat, na_rm = TRUE)
    , lon = first(lon, na_rm = TRUE)
    , addr = first(addr, na_rm = TRUE)
    ) %>%
  ungroup() %>%
  mutate(school_year = as.numeric(str_sub(school_year, start = 1, end = 4))) %>%
  mutate(across(-c(id, name, addr), as.numeric))


rm(df1, df2, df3, df4, df5)

tot_enroll <-
  df %>%
  group_by(school_year) %>%
  summarise(
      Total                 = sum(`Total Students All Grades (Excludes AE)`, na.rm=T)
    , `Free Lunch Eligible` = sum(`Free Lunch Eligible`, na.rm=T)
    , `Lower`               = sum(`Grades 1-8 Students`, na.rm=T)
    , `Upper`               = sum(`Grades 9-12 Students`, na.rm=T)
    , `Black`               = sum(`Black or African American Students`, na.rm=T)
    , `White`               = sum(`White Students`, na.rm=T)
    , `Hispanic`            = sum(`Hispanic Students`, na.rm=T)
    , `AAPI`                = sum(`Asian or Asian/Pacific Islander Students`, na.rm=T)
    , `Two or more`         = sum(`Two or More Races Students`, na.rm=T)
    , `AIAN`                = sum(`American Indian/Alaska Native Students`, na.rm=T)
    , Other                 = AAPI + AIAN + `Two or more`
    , tot_race              = sum(`Total Race/Ethnicity`, na.rm=T)
    , my_sum                = Black + Hispanic + Other + White
    , diff                  = round((my_sum - tot_race)/ tot_race*100, 2)
    , check_sum             = my_sum == tot_race
    , .groups               = 'drop'
  ) %>%
  mutate(across(where(is.numeric), ~na_if(., 0)))
```

<br>

# School enrollment data

DC's school enrollment data mirrors these Census trends.

Total enrollment in DC Public Schools (including charter schools) has risen from roughly `r fmt(tot_enroll$Total[tot_enroll$school_year == 1990])` to `r fmt(tot_enroll$Total[tot_enroll$school_year == 2023])` since 1990.^[U.S. Department of Education, National Center for Education Statistics' [Common Core of Data](https://nces.ed.gov/ccd/elsi/tableGenerator.aspx)] All of this increase came after 2010. Black student enrollment fell, while Hispanic and White student enrollment increased significantly.^[Starting around 2010, the Department of Education started listing "2 or more races as an option"; this altered totals for the other groups.]

<br>


```{r}
plt1 <-
  ggplot(tot_enroll %>% filter(school_year>=1990), aes(x=school_year)) +
  geom_line(aes(y=Total,    color="All"),      linewidth=ms) +
  geom_line(aes(y=Black,    color="Black"),    linewidth=ms) +
  geom_line(aes(y=Hispanic, color="Hispanic"), linewidth=ms) +
  geom_line(aes(y=Other,    color="Other"),    linewidth=ms) +
  geom_line(aes(y=White,    color="White"),    linewidth=ms) +
  # theme_minimal() +
  scale_y_continuous(
      labels = scales::label_comma()
    , limits = c(0, 100e3)
    ) +
  scale_color_manual(values = c(
              "All"=total_col
            , "Black"=black_col
            , "Hispanic"=hispa_col
            , "Other"=other_col
            , "White"=white_col
            )) +
  labs(x="", y="Enrollment", color = "",
       caption="Data includes charter schools. 'Other' includes students listed as AAPI, AIAN, and 2+ races.\nSum of groups may not sum exactly to 'All' category.\nSource: U.S. Department of Education, National Center for Education Statistics",
       title = "DCPS enrollment gains driven largely by Hispanic & White student enrollment"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    , panel.grid = element_blank()
    )


plt1
```

<br><br>


Since 1990, Wards 1, 7, and 8 have lost students, at least according to the DoE data that lists school locations. Wards 2, 3, 4, and 5 have seen significant increases. The trend in Ward 6 has been mostly flat.

<br>


```{r}
# Map schools
wards <- sf::st_read("data/Ward_-_2022.geojson", quiet=T)

points_sf <- df %>% 
  select(name, id, lat, lon) %>% distinct() %>%
  filter(!is.na(lat)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

joined_sf <- st_join(points_sf, wards, join = st_intersects) %>%
  select(name, id, WARD) %>% distinct() %>% st_drop_geometry()

df <- df %>%
  dplyr::left_join(joined_sf, by=c("id", "name"))


ward_enroll <- df %>%
  group_by(WARD, school_year) %>%
  summarise(
      Total                 = sum(`Total Students All Grades (Excludes AE)`, na.rm=T)
    , `Free Lunch Eligible` = sum(`Free Lunch Eligible`, na.rm=T)
    , `Lower`               = sum(`Grades 1-8 Students`, na.rm=T)
    , `Upper`               = sum(`Grades 9-12 Students`, na.rm=T)
    , `Black`               = sum(`Black or African American Students`, na.rm=T)
    , `White`               = sum(`White Students`, na.rm=T)
    , `Hispanic`            = sum(`Hispanic Students`, na.rm=T)
    , `AAPI`                = sum(`Asian or Asian/Pacific Islander Students`, na.rm=T)
    , `Two or more`         = sum(`Two or More Races Students`, na.rm=T)
    , `AIAN`                = sum(`American Indian/Alaska Native Students`, na.rm=T)
    , Other                 = AAPI + AIAN + `Two or more`
    , tot_race              = sum(`Total Race/Ethnicity`, na.rm=T)
    , my_sum                = Black + Hispanic + Other + White
    , diff                  = round((my_sum - tot_race)/ tot_race*100, 2)
    , check_sum             = my_sum == tot_race
    , .groups               = 'drop'
  ) %>%
  mutate(across(where(is.numeric), ~na_if(., 0))) %>%
  filter(!(is.na(WARD)))


plt2 <-
  ggplot(ward_enroll, aes(x=school_year, y=Total)) +
  geom_line(linewidth=ms-.2) +
  scale_y_continuous(
      labels = scales::label_comma()
    , limits = c(3e3, 18e3)
    ) +
  facet_wrap(~WARD) +
  labs(x="", y="Enrollment", 
       caption="Data includes charter schools. Some schools that did not have location data are omitted. \nSource: U.S. Department of Education, National Center for Education Statistics",
       title = "Enrollment up recently accross all wards, except wards 1, 7, and 8"
         ) +
  scale_color_brewer(palette = "Set1") +
  theme(
     plot.caption = element_text(hjust = 0)
    , panel.grid = element_blank()
    )

plt2
```

<br>

The chart below shows how student populations have changed by Ward and demographic group. Hispanic and White enrollment have increased; Black enrollment has decreased.

<br>



```{r}
ward_enroll_long <- ward_enroll %>%
  select(WARD, school_year, Black, Hispanic, Other, White) %>%
  pivot_longer(cols = -c(WARD, school_year), names_to = "Group", values_to = "Enrollment")

plt3 <-
  ggplot(ward_enroll_long, aes(x=school_year, y=Enrollment/1000, color=Group)) +
  geom_line(linewidth=ms-.2) +
  scale_y_continuous(
      labels = scales::label_comma()
    ) +
  scale_color_manual(values = c(
              "Black"=black_col
            , "Hispanic"=hispa_col
            , "Other"=other_col
            , "White"=white_col
            )) +
  facet_wrap(~WARD, scales="free_y") +
  labs(x="", y="Enrollment (in thousands)", color = "Ward",
       caption="Data includes charter schools. Some schools that did not have location data are omitted. \nSource: U.S. Department of Education, National Center for Education Statistics",
       title = "Enrollment up recently accross all wards, except wards 1, 7, and 8"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    , panel.grid = element_blank()
    )

plt3
```


<br><br>


# The reversal of white flight

Part of what is likely happening here is that white parents are more likely to move to the District (or remain in DC, if they were existing residents) than they were 40 years ago. This would be part of a broader trend: the reversal of white flight in large U.S. urban areas. Meanwhile, higher housing costs have led (or financially forced) many Black residents to move to the suburbs.^[See Yesim Sayin, Kids in the Neighborhood: The District Has More Children, But They Are Not Where They Used to Be. https://ora-cfo.dc.gov/blog/kids-neighborhood-district-has-more-children-they-are-not-where-they-used-be]


```{r}
races <- c(A = "White alone", B = "Black alone", C = "AIAN alone",
           D = "Asian alone", E = "NHPI alone", F = "Some other race alone",
           G = "Two or more races", H = "Hispanic or Latino",
           I = "White alone, not Hispanic")

spec <- tibble::tribble(
  ~year, ~sumfile, ~stem,
   2000,   "sf1",  "P034",
   2010,   "sf1",  "P038",
   2020,   "dhc",  "PCT10")

get_one <- function(year, sumfile, stem, geography = "state", state = "DC") {
  fmt <- if (year == 2020) \(s, i) sprintf("%s%s_%03dN", stem, s, i)
         else              \(s, i) sprintf("%s%s%03d",   stem, s, i)
  imap_dfr(races, function(lbl, sfx) {
    vars <- setNames(fmt(sfx, 1:20), sprintf("c%02d", 1:20))
    get_decennial(geography = geography, state = state, variables = vars,
                  year = year, sumfile = sumfile, output = "wide", key = CENSUS_KEY) |>
      mutate(race_hh = lbl, year = year)
  })
}

raw <- pmap_dfr(spec, get_one)

out <- raw |>
  transmute(year, GEOID, NAME, race_hh,
    families       = c01,
    any_child_u18  = c03 + c10 + c16,
    any_child_u6   = c04 + c05 + c11 + c12 + c17 + c18,
    any_child_6_17 = c05 + c06 + c12 + c13 + c18 + c19) |>
  arrange(year, race_hh)

a <-
  out %>%
  filter(race_hh %in% c("Black alone", "Hispanic or Latino", "White alone")) %>%
  ggplot(aes(x=race_hh, y=any_child_u6, fill=race_hh, group=factor(year))) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values = c(
              "Black alone"=black_col
            , "Hispanic or Latino"=hispa_col
            , "White alone"=white_col
            )) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x="", y="", 
       caption="Source: 2000, 2010, and 2020 Decennial Census."
       # , title    = "Fewer Black families, more White families with young children"
       , subtitle = "With children under 6"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    , panel.grid = element_blank()
    ) +
  guides(fill="none")

b <-
  out %>%
  filter(race_hh %in% c("Black alone", "Hispanic or Latino", "White alone")) %>%
  ggplot(aes(x=race_hh, y=any_child_6_17, fill=race_hh, group=factor(year))) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values = c(
              "Black alone"=black_col
            , "Hispanic or Latino"=hispa_col
            , "White alone"=white_col
            )) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x="", y=""
       , subtitle    = "With children 6-17"
       # , subtitle = "Number of families with children under 6 in the 2000, 2010, and 2020 Censuses, by race"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    , panel.grid = element_blank()
    ) +
  guides(fill="none")

a + b + plot_annotation(
  title = 'Fewer Black families, more White families with young children',
  subtitle = 'Number of families with children in the 2000, 2010, and 2020 Censuses',
  theme = theme(plot.title = element_text(size = 16, face = 'bold'))
)
```

<br>

We can even see this reversal if look at popular baby names registered with the Social Security Administration. First names more commonly associated with white people in the Census have jumped up, after falling over the latter half of the 20th century.^[Kaplan, J (2023). predictrace: Predict the Race and Gender of a Given Name Using Census and Social Security Administration Data. Version 2.0.1. URL: https://github.com/jacobkap/predictrace, https://jacobkap.github.io/predictrace/.]^[Note that for some names, like "Aiden" or "Genesis" the `predictrace` package doesn't return results.] 

<br>

```{r}
library(predictrace)
library(readxl)
library(zoo)

# Data comes from here:
# SSA baby names dataset for Washington DC
# https://www.ssa.gov/cgi-bin/namesbystate.cgi

names <- read_csv("data/DC.TXT", col_names = FALSE)
names(names) <- c("state", "gender", "year", "name", "count")

names$prob_w <- predictrace::predict_race(name = names$name, surname = F)$probability_white
names$prob_b <- predictrace::predict_race(name = names$name, surname = F)$probability_black

names %>% group_by(year) %>% summarise(mean = mean(prob_w, na.rm=T)) %>%
  ggplot(aes(x=year, y=mean)) +
  geom_point(color="steelblue", alpha=.8) +
  geom_line(aes(y = rollmeanr(mean, 5, na.pad = T)), color = "black", linewidth=.75) +
  theme_minimal() +
  scale_y_continuous(labels=scales::label_percent()) +
  labs(x="", y="Avg. modeled probability of 'white' name", color = "",
       caption="Source: Social Security Administration Popular Baby Names",
       title = "DC baby names track history of white flight (and return)"
         ) +
  theme(
     plot.caption = element_text(hjust = 0)
    , panel.grid = element_blank()
    )
```


# Income of households with children

```{r, results = 'hide', fig.keep = 'none', message = FALSE, warning = FALSE}
years <- c(2007, 2024)

# ── 1. Area median household income, each survey year ────────────────────
dc_msa_ami <- data.frame(
  year=c(2007, 2024),
  dc_msa_ami = c(94500, 154700) 
)

# print(natl_median)   # sanity-check these before going further


# ── 2. Pull DC person-level PUMS with replicate weights ──────────────────────
get_dc_year <- function(yr) {
  get_pums(
    variables   = c("AGEP", "HINCP", "NP"),
    state       = "DC",
    survey      = "acs1",
    year        = yr,
    rep_weights = "person",
    recode      = FALSE,
    key         = CENSUS_KEY
  ) %>%
    mutate(year = yr)
}

pums_raw <- map(years, get_dc_year) %>% set_names(years)

# Confirm replicate weights came back for BOTH years
# map(pums_raw, ~ sum(grepl("^PWGTP\\d+$", names(.x))))   # expect 80 each

# ── 3. Income bins relative to that year's national median ───────────────────
brks <- c(-Inf, 0.5, 0.8, 1.2, 2.0, Inf)
labs <- c("<50%", "50–80%", "80–120%", "120–200%", "200%+")

prep <- function(df, dc_msa_ami) {
  df %>%
    mutate(
      hincp     = as.numeric(HINCP),
      inc_ratio = hincp / dc_msa_ami,
      inc_grp   = cut(inc_ratio, breaks = brks, labels = labs,
                      right = FALSE, ordered_result = TRUE)
    )
}

# ── 4. Design first, then subset ─────────────────────────────────────────────
estimate_year <- function(yr) {
  ami <- dc_msa_ami$dc_msa_ami[dc_msa_ami$year == yr]

  svy <- pums_raw[[as.character(yr)]] %>%
    prep(dc_msa_ami = ami) %>%
    to_survey(type = "person", design = "rep_weights")

  svy %>%
    filter(AGEP < 18, !is.na(hincp)) %>%
    group_by(inc_grp) %>%
    summarise(
      children = survey_total(vartype = "se"),
      share    = survey_prop(vartype  = "se"),
      n_unwtd  = unweighted(n())
    ) %>%
    mutate(year = yr, dc_msa_ami = ami)
}

results <- map_dfr(years, estimate_year) %>%
  select(year, dc_msa_ami, inc_grp, children, children_se,
         share, share_se, n_unwtd)

# print(results, n = Inf)

a <- ggplot(results, aes(x=factor(year), fill=factor(inc_grp))) +
  geom_bar(aes(y=children), stat="identity", position='dodge') +
  labs(x="", y="Number of children") +
  guides(fill="none")

b <- ggplot(results, aes(x=factor(year), fill=factor(inc_grp))) +
  geom_bar(aes(y=share), stat="identity", position='dodge') +
  labs(x="", fill="Income rel. to AMI") +
  theme(legend.position = "bottom")
```

```{r}
(a + b) +
  plot_annotation(title = "Largest % growth in children from higher-income households") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```


<br>
<br>
<br>


# Marriage, birth, and fertility rate trends

```{r}
# Data from here:
# https://www.bgsu.edu/ncfmr/resources/data/original-data/county-level-birth-data-1982-2024
df <- read_excel("data/ncfmr-dc-cntyyr-st-ntl-long-births-rates-1982-2024.xlsx")

df <- dplyr::inner_join(
    out_c %>% select(GEOID, NAME, city) %>% distinct()
  , df, by = c("GEOID"="geoid"))
```


The chart below shows the number of births in DC over time. Births in DC increased modestly during the 2010s, then fell again.^[See also https://edscape.dc.gov/page/pop-and-students-number-births-and-birth-rates]

<br>

```{r}
ggplot(df %>% filter(city=="Washington")
       , aes(x=year, y=births)) +
  geom_line(linewidth=ms) +
  scale_y_continuous(limits = c(0, 12e3)) +
  labs(x="", y="Births in DC", 
       title="The number of births in DC has risen only slightly since 2000")
```


<br>

Demographers calculate the "general fertility rate" by dividing the births by the female population ages 15 to 44, and multiplying by 1,000.^[https://www.ncbi.nlm.nih.gov/books/NBK215693/] The chart below shows DC's GFR over time. This has fallen continuously. 

This suggests that the increasing number of births in DC was less driven by a higher percentage of residents deciding to have children, and more driven by:

1. Increased in-migration of people who would go on to have children in DC
2. Decreased out-migration of people who would go on to have children in DC

<br>

```{r}
ggplot(df %>% filter(city=="Washington")
       , aes(x=year, y=gfr)) +
  geom_line(linewidth=ms) +
  scale_y_continuous(limits = c(0, 75)) +
  labs(x="", y="General Fertility Rate", 
       title="DC's GFR has fallen over time")
```

<br>

DC's GFR is on the lower end among large U.S. cities.

<br>


```{r}
df %>% filter(year==2024) %>%
  group_by(city, year) %>%
  summarise(
      births=sum(births, na.rm=T)
    , age1544_fem=sum(age1544_fem, na.rm=T)
    , gfr = births / age1544_fem
    , .groups='drop'
    ) %>%
  mutate(DC = if_else(city == "Washington", 1, 0)) %>%
  ggplot(aes(y=forcats::fct_reorder(city, gfr), x=gfr*1000, fill=factor(DC))) +
  geom_bar(stat="identity") +
  labs(x="GFR in 2024", y="") + 
  guides(fill="none") +
  # scale_x_continuous(labels = scales::label_comma()) +
  labs(title="")

```

<br>

The chart below shows general fertility rates for the counties encompassing different U.S. cities between 1982 and 2024. The vast majority of U.S. cities saw large declines in the GFR. DC's decline was sort of in the middle.

<br>

```{r}
  df %>% 
  group_by(city, year) %>%
  summarise(
      births=sum(births, na.rm=T)
    , age1544_fem=sum(age1544_fem, na.rm=T)
    , gfr = births / age1544_fem
    , .groups='drop'
    ) %>%
  group_by(city) %>%
  mutate(change = gfr[year==1985] - gfr[year==2024]) %>%
  ungroup() %>%
  mutate(DC = if_else(city == "Washington", 1, 0)) %>%
  ggplot(aes(x=year, y=gfr
             , color=factor(DC)
             )
         ) +
  geom_line() +
  labs(x="", y="GFR") +
  guides(color="none") +
  scale_color_manual(values = c("black", "salmon")) +
  theme(
    panel.grid = element_blank(),
    axis.title.y = element_blank(), # Removes the main Y-axis title ("mpg")
    axis.text.y  = element_blank(), # Removes the tick labels (numbers)
    axis.ticks.y = element_blank(), # Removes the small tick marks
    
    axis.ticks.x = element_blank(),  # Removes the small tick marks
    axis.text.x  = element_blank()  # Removes the tick labels (numbers)
  ) +
  facet_wrap(~forcats::fct_reorder(city, change), scales="free_y")
```


```{r}
# =========================================================================
# 1. 2020 CENSUS DATA (DHC File)
# =========================================================================
vars_2020 <- c(
  # Males (Ages 0 to 18)
  "M0" = "PCT12_003N", "M1" = "PCT12_004N", "M2" = "PCT12_005N", "M3" = "PCT12_006N", 
  "M4" = "PCT12_007N", "M5" = "PCT12_008N", "M6" = "PCT12_009N", "M7" = "PCT12_010N", 
  "M8" = "PCT12_011N", "M9" = "PCT12_012N", "M10" = "PCT12_013N", "M11" = "PCT12_014N", 
  "M12" = "PCT12_015N", "M13" = "PCT12_016N", "M14" = "PCT12_017N", "M15" = "PCT12_018N", 
  "M16" = "PCT12_019N", "M17" = "PCT12_020N", "M18" = "PCT12_021N",
  # Females (Ages 0 to 18)
  "F0" = "PCT12_107N", "F1" = "PCT12_108N", "F2" = "PCT12_109N", "F3" = "PCT12_110N", 
  "F4" = "PCT12_111N", "F5" = "PCT12_112N", "F6" = "PCT12_113N", "F7" = "PCT12_114N", 
  "F8" = "PCT12_115N", "F9" = "PCT12_116N", "F10" = "PCT12_117N", "F11" = "PCT12_118N", 
  "F12" = "PCT12_119N", "F13" = "PCT12_120N", "F14" = "PCT12_121N", "F15" = "PCT12_122N", 
  "F16" = "PCT12_123N", "F17" = "PCT12_124N", "F18" = "PCT12_125N"
)

dc_2020_raw <- get_decennial(
  geography = "state", state = "DC", key=CENSUS_KEY,
  variables = vars_2020, year = 2020, sumfile = "dhc"
)

dc_2020_binned <- dc_2020_raw %>%
  mutate(age = as.numeric(str_remove(variable, "^[MF]"))) %>%
  mutate(age_group = case_when(
    age < 3  ~ "under 3",         age < 5  ~ "3 but under 5",
    age < 7  ~ "5 but under 7",   age < 9  ~ "7 but under 9",
    age < 11 ~ "9 but under 11",  age < 13 ~ "11 but under 13",
    age < 15 ~ "13 but under 15", age < 17 ~ "15 but under 17",
    age < 19 ~ "17 and under 19"
  )) %>%
  group_by(age_group) %>%
  summarise(pop_2020 = sum(value))

# =========================================================================
# 2. 2010 CENSUS DATA (SF1 File)
# =========================================================================
vars_2010 <- c(
  # Males (Ages 0 to 18)
  "M0" = "P014003", "M1" = "P014004", "M2" = "P014005", "M3" = "P014006", 
  "M4" = "P014007", "M5" = "P014008", "M6" = "P014009", "M7" = "P014010", 
  "M8" = "P014011", "M9" = "P014012", "M10" = "P014013", "M11" = "P014014", 
  "M12" = "P014015", "M13" = "P014016", "M14" = "P014017", "M15" = "P014018", 
  "M16" = "P014019", "M17" = "P014020", "M18" = "P014021",
  # Females (Ages 0 to 18)
  "F0" = "P014024", "F1" = "P014025", "F2" = "P014026", "F3" = "P014027", 
  "F4" = "P014028", "F5" = "P014029", "F6" = "P014030", "F7" = "P014031", 
  "F8" = "P014032", "F9" = "P014033", "F10" = "P014034", "F11" = "P014035", 
  "F12" = "P014036", "F13" = "P014037", "F14" = "P014038", "F15" = "P014039", 
  "F16" = "P014040", "F17" = "P014041", "F18" = "P014042"
)

dc_2010_raw <- get_decennial(
  geography = "state", state = "DC", key=CENSUS_KEY,
  variables = vars_2010, year = 2010, sumfile = "sf1"
)

dc_2010_binned <- dc_2010_raw %>%
  mutate(age = as.numeric(str_remove(variable, "^[MF]"))) %>%
  mutate(age_group = case_when(
    age < 3  ~ "under 3",         age < 5  ~ "3 but under 5",
    age < 7  ~ "5 but under 7",   age < 9  ~ "7 but under 9",
    age < 11 ~ "9 but under 11",  age < 13 ~ "11 but under 13",
    age < 15 ~ "13 but under 15", age < 17 ~ "15 but under 17",
    age < 19 ~ "17 and under 19"
  )) %>%
  group_by(age_group) %>%
  summarise(pop_2010 = sum(value))

# =========================================================================
# 3. 2000 CENSUS DATA (SF1 File)
# =========================================================================
vars_2000 <- c(
  # Males (Ages 0 to 18)
  "M0" = "PCT012003", "M1" = "PCT012004", "M2" = "PCT012005", "M3" = "PCT012006", 
  "M4" = "PCT012007", "M5" = "PCT012008", "M6" = "PCT012009", "M7" = "PCT012010", 
  "M8" = "PCT012011", "M9" = "PCT012012", "M10" = "PCT012013", "M11" = "PCT012014", 
  "M12" = "PCT012015", "M13" = "PCT012016", "M14" = "PCT012017", "M15" = "PCT012018", 
  "M16" = "PCT012019", "M17" = "PCT012020", "M18" = "PCT012021",
  # Females (Ages 0 to 18)
  "F0" = "PCT012107", "F1" = "PCT012108", "F2" = "PCT012109", "F3" = "PCT012110", 
  "F4" = "PCT012111", "F5" = "PCT012112", "F6" = "PCT012113", "F7" = "PCT012114", 
  "F8" = "PCT012115", "F9" = "PCT012116", "F10" = "PCT012117", "F11" = "PCT012118", 
  "F12" = "PCT012119", "F13" = "PCT012120", "F14" = "PCT012121", "F15" = "PCT012122", 
  "F16" = "PCT012123", "F17" = "PCT012124", "F18" = "PCT012125"
)

dc_2000_raw <- get_decennial(
  geography = "state", state = "DC", key=CENSUS_KEY,
  variables = vars_2000, year = 2000, sumfile = "sf1"
)

dc_2000_binned <- dc_2000_raw %>%
  mutate(age = as.numeric(str_remove(variable, "^[MF]"))) %>%
  mutate(age_group = case_when(
    age < 3  ~ "under 3",         age < 5  ~ "3 but under 5",
    age < 7  ~ "5 but under 7",   age < 9  ~ "7 but under 9",
    age < 11 ~ "9 but under 11",  age < 13 ~ "11 but under 13",
    age < 15 ~ "13 but under 15", age < 17 ~ "15 but under 17",
    age < 19 ~ "17 and under 19"
  )) %>%
  group_by(age_group) %>%
  summarise(pop_2000 = sum(value))

# =========================================================================
# 4. COMBINE AND SORT ALL YEARS
# =========================================================================
dc_final_trends <- dc_2000_binned %>%
  full_join(dc_2010_binned, by = "age_group") %>%
  full_join(dc_2020_binned, by = "age_group") %>%
  mutate(age_group = factor(age_group, levels = c(
    "under 3", "3 but under 5", "5 but under 7", "7 but under 9", 
    "9 but under 11", "11 but under 13", "13 but under 15", 
    "15 but under 17", "17 and under 19"
  ))) %>%
  filter(age_group != "17 and under 19") %>%
  arrange(age_group) %>%
  mutate(
      pct_2000 = pop_2000 / sum(pop_2000)
    , pct_2010 = pop_2010 / sum(pop_2010)
    , pct_2020 = pop_2020 / sum(pop_2020)
    , change_pct_2000_2020 = pct_2020 - pct_2000
    , raw_change = pop_2020 - pop_2000
    )
```

<br>

The table below shows population trends by age group. The population of very young children has grown, while other age groups' populations have declined somewhat (although 2020 may have been an especially low year).

<br>

```{r}
dc_final_trends %>%
  select(-starts_with("pct_"), -change_pct_2000_2020) %>%
  rename(  `Age group` = age_group
         , `2000 population` = pop_2000
         , `2010 population` = pop_2010
         , `2020 population` = pop_2020
         , `'00 - '20 change` = raw_change
         ) %>%
  mutate(`Tot. change '00 - '20` = sum(`'00 - '20 change`)) %>%
  knitr::kable()
```

<br>

Net migration for children in DC has been negative for decades, but during the 2010s, it was less negative.

<br>

```{r}
temp_df <- read_csv("~/GitHub/youth-stats/data/net_migration_wisc.csv") %>%
  mutate(group = case_when(
      group=="Total"~"All"
    , group=="NH_Black" ~ "Non-Hisp. Black"
    , group=="NH_White" ~ "Non-Hisp. White" 
    )
    ) %>%
  mutate(age = factor(age, levels = c("0-4", "5-9", "10-14", "15-19"))) %>%
  filter(age != "15-19")

ggplot(temp_df, aes(x=age, y=net_migration_per_100_people, fill=factor(decade))) +
  geom_bar(stat = "identity", position="dodge") +
  scale_fill_manual(values = c("#041C39", "#1070E5", "#67A6F4")) +
  facet_grid(~group) +
  labs(
      x="", y="Net migration per 100 people", title="Child net migration increased in the 2010s"
      , fill=""
    , caption="Egan-Robertson, David, Katherine J. Curtis, Richelle L. Winkler, Kenneth M. Johnson, and Caitlin Bourbeau,\nAge-Specific Net Migration Estimates for US Counties, 1950-2020.\nApplied Population Laboratory, University of Wisconsin - Madison, 2024. Web.") +
  theme(
    plot.caption = element_text(hjust = 0)
  )
```

<br>

This tracks with population trends across all age groups, which show that DC's 2010s population boom was largely driven by increasing net migration.

<br>

```{r}
df <- read_csv("~/GitHub/youth-stats/data/components_of_population_change.csv") %>%
  select(-`...11`) %>%
  select(Description, Year, Births, Deaths
         , `Net International Migration`, `Net Domestic Migration`, `Residual`) %>%
  mutate(Deaths = Deaths * -1) %>%
  pivot_longer(cols = -c("Description", "Year"), names_to = "var") %>%
  filter(Description %in% c(
      "District of Columbia"
    , "Arlington County, VA"
    , "Montgomery County, MD"
    , "Prince George's County, MD"
    )
    ) %>%
  filter(Year!=1999) %>%
  group_by(Year, Description) %>%
  mutate(Total = sum(value)) %>%
  ungroup() %>%
  mutate(var = case_when(
      var=="Net Domestic Migration" ~ "Net Dom. Mig."
    , var=="Net International Migration" ~ "Net Intl. Mig."
    , .default=var
  ))

ggplot(data=df) +
  geom_bar(
             aes(x=Year, y=value, fill=factor(var))
           , stat="identity"
           ) +
  geom_line(aes(x=Year, y=Total)) +
  facet_wrap(~Description) +
  labs(x="", y="Population gain/loss", fill=""
       , caption="Source: StatsAmerica, U.S. Census Bureau"
       , title="DC's 2010s population growth driven by net migration"
       ) +
  theme(
    plot.caption = element_text(hjust=0)
  )
```

<br>

Finally, the chart below shows marriage and divorce counts.^[National Center for Family & Marriage Research (2023). County-level marriage and divorce data, United States, 2011-2019, (Version 1.0). Marriage, Divorce, & Birth Data Compass. Bowling Green State University. Retrieved from http://www.bgsu.edu/ncfmr/resources/data/original-data/county-level-marriage-divorce-data-2011-2019.html ]

<br>


```{r}
# Source: https://www.bgsu.edu/ncfmr/resources/data/original-data/county-level-marriage-divorce-data-2011-2019.html
df <- read_excel("data/CL-Mar-Div-2011-2019.xlsx") %>%
  select(STATE, COUNTY, GISJOIN, starts_with("M"), starts_with("D")) %>%
  pivot_longer(cols = -c(STATE, COUNTY, GISJOIN)
               , names_to = c(".value", "year")
               , names_pattern = "^(.*)([0-9]{4})$"
               ) %>%
  mutate(year = as.numeric(year))

ggplot(df %>% filter(COUNTY == "District of Columbia"), aes(x=year)) +
  geom_line(linewidth=ms, aes(y=M, color="M")) +
  geom_line(linewidth=ms, aes(y=D, color="D")) +
  scale_y_continuous(limits=c(0, 8000)) +
  scale_x_continuous(n.breaks = 10) +
  labs(x="", y="Number of marriages/divorces", 
       title="DC's number of marriages and divorces relatively stable in the 2010s") +
  scale_color_manual(
    name = "",                     # Legend title
    labels = c("Divorces", "Marriages"),      # Optional custom labels
    values = c("D"="grey", "M"="steelblue")
    )
```


```{r}

temp <-
  df %>% filter(GISJOIN %in% c("G1100010", "G2500250", "G2405100", "G4201010", "G0600750")) %>%
  group_by(GISJOIN) %>%
  summarise(M = mean(M, na.rm=T), D = mean(D, na.rm=T), .groups="drop") %>%
  mutate(city = case_when(
      GISJOIN == "G0600750" ~ "San Francisco"
    , GISJOIN == "G4201010" ~ "Philadelphia"
    , GISJOIN == "G2405100" ~ "Baltimore"
    , GISJOIN == "G2500250" ~ "Boston"
    , GISJOIN == "G1100010" ~ "Washington"
  )) %>%
  dplyr::left_join(select(out_c, year, city, population, race) %>% 
                     filter(year==2020, race == "Total, all races") %>% 
                     select(-year, -race) %>% distinct()
                   , by="city") %>%
  mutate(m_rate = M / population * 1000, d_rate = D / population * 1000)
  
a <- ggplot(temp, aes(x=city, y=m_rate)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits=c(0, 15)) +
  labs(x="", y="Marriages per 1000 residents", 
       title="DC has relatively few marriages per 1000 residents") 

b <- ggplot(temp, aes(x=city, y=d_rate)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits=c(0, 10)) +
  labs(x="", y="Divorces per 1000 residents", 
       title="... and more divorces") 

a / b
```

<br>
<br>
<br>

# Conclusion

* Some of DC's increase in youth-age population during the 2010s was due to an increased number of births, because more people giving birth lived here--but it was not do to an increase in the fertility rate of the population. 
* A lot of this increase in the number of births was due to increased net migration of families with children. 
* This increased net migration seemed to be concentrated among Hispanic, White, and higher-income residents. DC's population of Black children has fallen over time. While the number of low-income children has grown, the share of children in DC from lower-income families has shrunk.
* Some of DC's youth-age population growth is, statistically, a reversion to levels seen during the 1990s.




